---
title: "The Value of Life: A Macro Investigation of GDP and Life Expectancy Correlation"
author: 
  - Abigail Tata^[American University]
  - Kenneth Moreno^[American University]
  - Joshua Greene^[American University]
  - Fabiola Cruz^[American University]
format: 
  pdf:
    number-sections: true
    toc: false
    geometry: 
      - top=3cm
date: 2024-01-12
date-format: iso
documentclass: article
editor: source
abstract: "This paper seeks to investigate the relationship between GPD per capita and life expectancy rates worldwide. We do so by using xxx functions and xxx in R with data from the World Bank for year 2015 to help establish a statistically significant relationship between the dependent variable (life expectancy) and the independent variable (GDP per capita) - we also include additional variables for control. The results  show that GDP per capita is not a statistically significant predictor of life expectancy, instead, birth rates and death rates are more effective predictors of life expectancy. Additionally, we also reaffirm that countries with higher birth rates tend to have a lower life expectancy"
bibliography: main.bib

---

```{r setup}
#| echo: false
#| message: false

library(Sleuth3)
library(tidyverse)
library(car)
library (mosaic)
library (ggthemes)
library (gmodels)
library (DescTools)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidymodels)
library(tidyr)
library(modelr)

```

```{r}
#| eval: false
#| echo: false
#| message: false
LE1<- read_csv("countrydata.csv")
# read in income class data 
incomeclass<- read_csv("incomeclass.csv")
view(incomeclass)

# take out NAs in the dataset 
LE1 <- na.omit(LE1)
sum(is.na(LE1$GDP.per.capita))
sum(is.na(LE1$Current.health.expenditure.per.capita))


# make all variables numeric 
LE1 <- LE1 %>%
  mutate(
    GDP.per.capita = as.numeric(as.character(GDP.per.capita)),
    Population = as.numeric(as.character(Population )),
    Life.expectancy = as.numeric(as.character(Life.expectancy)),
    Current.health.expenditure.per.gdp = as.numeric(as.character(Current.health.expenditure.per.gdp)),
    Current.health.expenditure.per.capita = as.numeric(as.character(Current.health.expenditure.per.capita)),
    Compulsory.education.duration = as.numeric(as.character(Compulsory.education.duration )),
    Birth.rate.p1000  = as.numeric(as.character(Birth.rate.p1000 )),
    Death.rate.p1000 = as.numeric(as.character(Death.rate.p1000 ))
)


# log

# log  GDP
LE1$logGDP<- log(LE1$GDP.per.capita)
# log Current health expenditure per capita
LE1$logHCExpCapita<- log(LE1$Current.health.expenditure.per.capita)

```

```{r readdata}
#| eval: false
#| echo: false
#| message: false


merged_df <- merge(incomeclass, LE1, by = "Country.Name")
```

# Introduction

One of the best indicators to measure socio-economic stability in a country is life expectancy. It not only helps us understand how countries are effectively providing basic services to citizens, but serves as a clue to the country’s overall governance infrastructure which can determine access to health care, jobs, education, crime rates, and much more. In addition, to fund the social services that a society needs to thrive, the country also needs to ensure a robust and thriving economy. And even with a robust economy, sometimes disparities in quality of life and access to services can increase. In the United States for example, @national_academies_of_sciences_growing_2015 illustrates how the gap between income and life expectancy has widened over recent decades. Thus, we seek to ask: Does a higher GDP per capita have an impact on life expectancy?

We hypothesize that higher GDP per capita will indeed have a direct effect on life expectancy of a country, which is exemplified in US social phenomenons like the access “nice” neighborhoods have to healthcare. Using data and statistical software, we aim to substantiate our claims that GDP per capita and life expectancy are directly correlated. We examined this  relationship by pulling data on life expectancy and GDP per capita from the World Development Indicators [-@world_bank_world_2015], tidying the data, and running **xxx** and **xxx** in R. When we measured for GDP per capita and life expectancy alone, we found a strong correlation between the two. While this initially proved our hypothesis, when we accounted for factors like compulsory education and health expenditures, that initially strong correlation was weak. However, we were able to confirm that birth rates and death rates are better predictors of life expectancy, finding them statistically significant to help us identify a correlation. Finally, we also were able to determine that countries with higher birth rates tend to have lower life expectancy rates - a metric that made us think further about the socio-economic forces determining said factors.  A more detailed analysis of the theoretical evidence, the data, our methods, and the results are outlined below. 

# Framing the Conversation

Scholars have sought to understand how the economic status of nations around the world correlates with the health of their citizens since the late 1900s. Low-income nations often experience the highest mortality rates among their population, a phenomenon our research seeks to attribute to economic performance. Ultimately, studies done on the health outcomes of nations are often bogged down by the governance and statistical capacity deficits. Health issues are multidimensional, often incorporating social and economic factors that are hard to measure. However, the research is incredibly important to fully understand wealth and life expectancy disparities in society, in hopes that it leads us towards sustainable solutions. Through a deeper understanding of the ways in which these factors influence each other, we can better understand why some countries remain economically stagnant and why others prosper. 

While medicinal innovation has been booming since the late 1940s with the creation of organizations like the  World Health Organization (WHO), many countries lack the means to access lifesaving care. @acemoglu_disease_2007 argue that this inaccessibility is directly related to their life expectancy rates, stating that the lack of a sustained workforce inhibits economic production and participation, butting low-income countries out of the fighting ring for health resources. They also argue that the temporary supply of capital to nations with low GDP per capitas boosts their health outcomes, marginally, but failure to create a sustainable economy drops them back down after a while. Low-income nations also routinely spend more time and resources on fertility, growing their population to make up for high mortality rates. However, @simona_relationship_2014 argues that this practice has the opposite effect and instead produces a population that the nation’s economy routinely fails to support thus leading to further economic and health strains. Such an unfocused effort towards improving mortality rates and life expectancy within the country is exactly why research into GDP per capita and life expectancy correlation is so vital. A clear, concise route to health improvement and economic prosperity starts with a base understanding of how these two factors routinely rely on each other. 

Geopolitical and domestic policies—and conflicts—can also have a strong effect on the GDP per capita of a nation and their subsequent life expectancy. As argued by @national_academies_of_sciences_growing_2015, a gap between income and life expectancy globally has major implications for economic sustainability, eliciting fears of a quickly decreasing rate of global production due to a concentration of resources exclusive to the highest-earning nations. In times of global or regional conflict, the material differences between developing and developed countries are exacerbated as well. Whether or not the country has the resources to mitigate the suffering of their citizens in times of political strife is directly related to their economic capability.

Unarguably, the implications of this research are vast. An informed understanding of the economic effects of mortality rates and overall health can lead to a more substantive response, helping ensure future economic stability and the upwards mobility of developing nations. Furthermore, it is important to note that this research does not only seek to draw focus on low-income nations but nations from middle to high-income as well. Measuring the relationship between GDP per capita and life expectancy has implications that go across the macro-economic spectrum. What follows is our collection and analysis of hard data acquired through the World Bank, which will serve to add to the conversation of the dependence health outcomes have on economic prosperity within a country–fostering what we hope to be a step towards a more sustainable solution to global health inequity. 

# Data

# Results

# Discussion

In undertaking this topic, we aimed to find and measure a correlation between GDP per capita and life expectancy around the world. Understanding how these two factors influence each other is the keystone to tackling significant global issues, like poverty and high mortality rates. We found that measured on their own, GDP per capita and life expectancy had a strong correlation: the higher the GDP per capita; the higher the life expectancy. We had predicted this outcome early on, attributing the relationship to a country’s access or lack of access to life-sustaining resources and quality of life. However, when accounting for factors beyond GDP per capita, like compulsory education and population size, we found that the relationship was weakened significantly. To further explore this, we also measured birth rates, death rates, and life expectancy and found a strong correlation between high birth rates, death rates and low life expectancy. This relationship helped inform our analysis and possible future implications. 

Throughout this research, a cyclical phenomenon became clear. Countries with low GDP per capita lacked the resources to sustain their population, thus leading to high mortality rates. As @simona_relationship_2014 argued, high mortality rates incentivized citizens to focus on fertility to “make up” for the deaths. This, however, adds strain to an economy that is already unable to sustain its current population. Thus, more of the population is left economically stratified in poor living conditions, heightening the risk of death to preventable illness or disease. An unsustainable population means an unsustainable workforce, leading to the limited production, innovation, and a lack of participation in the national economy, culminating in a low GDP per capita. This cycle is doomed to continue to repeat itself. 

Of course, this research was limited in some ways. As stated, we found that when we accounted for other factors, the correlation between GDP per capita and life expectancy was not as statistically significant as we had predicted. While surprising, this outcome did have a precedent in previously published research. As stated before, many factors that contribute to a country’s life expectancy are notoriously multidimensional and are difficult to measure. Ultimately, our topic focused on the factors we could accurately attest for – GDP per capita, death rates, birth rates, population size, and compulsory education – and while we discovered that GDP per capita was not as strong of a predictor for life expectancy as we had hoped, it was incredibly important supplemental data in understanding the correlation between birth rates, death rates, and life expectancy. 

Undoubtedly, this topic is broad and nuanced. In future replications of this research, one should be cognizant of the historical and geopolitical factors that contribute to a country’s life expectancy or GDP per capita. While we found relationships that are significant and important, and of course lead to deeper understandings of global problems, a holistic approach to this topic has the possibility of producing a meaningful framing of the issues plaguing low-income countries. It also has the potential to be incredibly beneficial in informing future international economic policy and the mitigation of risk for high-income countries. And while our research was limited in scope, we hope it contributes to a broader conversation on health and overall socio-economic equity.

\clearpage

# References
